Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
represented annotated proofs as files rather than showing them fully in tex

Logical atomicity is key to the template approach: each template's \lstinline{traverse} function is proved to satisfy a logically atomic specification that says roughly ``this function finds the node where key \lstinline{k} belongs''. The traverse specifications can then be used to prove atomic specifications for the individual data structure operations, lifting the sequential specifications for insert, lookup, etc. to the concurrent setting.

@@

Recall that in the lock-coupling pattern, each thread always holds at least one lock while traversing the data structure, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search. Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{n}$ and obtain the resources $\mathsf{R}(\texttt{n})$ guarded by the lock. These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the resource  $\mathsf{R}(\texttt{n}) \triangleq \mathsf{ghost\_node(\texttt{n}, k, v, range)} \ \ast \ \mathsf{DataStruct\_specific(\texttt{n})}$.

@@

In the specification, the local precondition includes both the node handle $\nodeboxrep(\texttt{n})$ and the contents $\mathsf{R}(\texttt{n})$. The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}. The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a Boolean variable $\mathit{res}$ indicating whether \texttt{n'->t} is \texttt{NULL} or not. The variable $\mathit{res}$ is $\mathsf{true}$ if $\texttt{flag}$ returns 1, and $\mathsf{false}$ if 0, mirroring its Boolean state. The key to the correctness of the \texttt{traverse} function the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range}(\texttt{n'})\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \  \mathsf{R}(\texttt{n'}) \ \ast \ \treerep\ m
\end{mathpar} %\wm{Where does \lstinline{range} come from here?}

@@

The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup},  is simpler than that of the lock-coupling template. The local precondition include any locks, and consists only of $\infp(\texttt{n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$. As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$. Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{n'->t} is \texttt{NULL} or not.

@@

As previously mentioned, the \texttt{insertOp} specification (Figure \ref{fig:insertOp_giveup}) is slightly different from the lock-coupling version, to account for the added \lstinline{min} and \lstinline{max} fields. In a BST, if the current node's range is \lstinline{(min, max)} and its key is \lstinline{k}, its left child will have range \texttt{(min, k)} and its right child will have range \texttt{(k, max)}. Otherwise, \lstinline{insertOp} behaves identically, storing \lstinline{k} and \lstinline{v} in the current node and creating two empty child nodes.

@@

	\forall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \infp (\texttt{n}) \ \big| \ \treerep\ m
		node_t *r = (pn->n);
		\infp (\texttt{n})  \ast \infp (\texttt{r}) \ast \treerep\ m 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
			\texttt{pn} \mapsto (\texttt{p, n}) \ast \infp (\texttt{n}) \ast \infp (\texttt{r}) \ast \treerep\ m
				$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots \end{array}\right\}$
				pn->p = pn->n; $\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots \end{array}\right\}$
				if (inRange(pn, k) == 1){
				$\color{blue} \left\{\begin{array}{l} \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast  \cdots \end{array}\right\}$
								$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} = \texttt{NULL}  \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$
								(\mathit{res'} = \mathsf{false} \ast \texttt{n'} = \texttt{n} \ast  \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots )\  \lor \\ 
								(\mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \infp (\texttt{n'}) \ast \cdots)   \end{array}\right\}$
										$\color{violet} \left\{\begin{array}{l} \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL}  \ast \texttt{n'} = \texttt{n} \ast  \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots   \end{array}\right\}$
								else
										$\color{blue} \left\{\begin{array}{l} \mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n->t} \neq \texttt{NULL} \ast \infp (\texttt{n'}) \ast \cdots    \end{array}\right\}$
										$\color{blue} \left\{\begin{array}{l} \mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n->t} \neq \texttt{NULL} \ast \infp (\texttt{n'}) \ast \cdots    \end{array}\right\}$
						$\color{blue} \left\{\begin{array}{l} \texttt{k} \not\in \mathsf{range}(\texttt{n}) \ast \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast  \cdots \end{array}\right\}$
						$\color{blue} \left\{\begin{array}{l} \texttt{k} \not\in \mathsf{range}(\texttt{n}) \ast \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{n, n}) \ast \infp (\texttt{n}) \ast  \cdots \end{array}\right\}$
						pn->n = r;
						$\color{blue} \left\{\begin{array}{l} \texttt{k} \not\in \mathsf{range}(\texttt{n}) \ast  \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{r, r}) \ast \infp (\texttt{r}) \ast \cdots    \end{array}\right\}$
		$\color{blue} \left\{\begin{array}{l} (\mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots) \  \lor \\
		(\mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots)\end{array}\right\}$
	\begin{array}{c}
		\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
		\\ \infp(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
	\caption{The \texttt{traverse} function in give-up template annotated with separation logic specification}
	\label{proof_giveup_traverse}

\begin{figure}[!h]
	$\color{blue}
	\forall m.\left\langle 
	\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
	\right\rangle$
	\begin{lstlisting}[language = C, style=myStyle, mathescape=true]
void insert (node_t **r, int k, void *v){
		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
		pn->n = *r;
		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{p, r}) \ast  \texttt{r->lock} \lockvar \mathsf{R}(\texttt{r}) \ast \treerep\ m \end{array}\right\}$
		acquire(pn->n->lock);
		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p, r}) \ast  \texttt{r->lock} \lockvar \mathsf{R}(\texttt{r}) \ast \mathsf{R}(\texttt{r}) \ast \cdots \end{array}\right\}$
		if (traverse(pn, k) == 0){
		$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL} \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
				pn->p->t->value = v;
				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \texttt{p'->t->value} \mapsto \texttt{v} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
				//Linearization point
				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
		}
		else{
				$\color{blue} \left\{\begin{array}{l}  \mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
				insertOp(pn, k, v);
				$\color{blue} \left\{\begin{array}{l}  \mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast  \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \\ \texttt{p'->t}\mapsto (\texttt{k, v', l, r})  \ast \texttt{l} \mapsto \texttt{NULL} \ast  \texttt{r} \mapsto \texttt{NULL} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
				//Linearization point
				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
		}
		release(pn->n->lock);
		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
		free(pn);
		$\color{blue} \left\{\begin{array}{l}  \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
} \end{lstlisting}
	$\color{blue}
	\left\langle 
	\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
	\right\rangle$
	\caption{The \texttt{insert} function in lock-coupling template annotated with separation logic specification}
	\label{proof_lock_insert}
\end{figure}

\begin{figure}[!h]
	$\color{blue}
	\forall m.\left\langle 
	\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
	\right\rangle$
	\begin{lstlisting}[language = C, style=myStyle, mathescape=true]
void insert (node_t **r, int k, void *v){
		struct pn *pn = (struct pn *) surely_malloc (sizeof *pn);
		pn->n = *r;
		$\color{blue} \left\{\begin{array}{l} \texttt{pn} \mapsto (\texttt{p, r}) \ast  \infp(\texttt{r}) \ast \treerep\ m \end{array}\right\}$
		if (traverse(pn, k) == 0){
				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
				pn->p->t->value = v;
				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \texttt{p'->t->value} \mapsto \texttt{v} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
				//Linearization point
				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
		}
		else{
				$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots \end{array}\right\}$
				insertOp(pn, k, v);
				$\color{blue} \left\{\begin{array}{l}  \mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast  \infp (\texttt{p'}) \ast \\ \texttt{p'->t}\mapsto (\texttt{k, v', l, r})  \ast \texttt{l} \mapsto \texttt{NULL} \ast  \texttt{r} \mapsto \texttt{NULL} \ast \cdots \end{array}\right\} \Rrightarrow{\textbf{commit}}$ 
				//Linearization point
				$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \cdots \end{array}\right\}$
		}
		release(pn->n->lock);
		$\color{blue} \left\{\begin{array}{l}  \texttt{pn} \mapsto (\texttt{p', p'}) \ast \infp (\texttt{p'}) \ast \cdots \end{array}\right\}$
		free(pn);
		$\color{blue} \left\{\begin{array}{l} \infp (\texttt{p'}) \ast \cdots \end{array}\right\}$
	} \end{lstlisting}
	$\color{blue}
	\left\langle 
	\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
	\right\rangle$
	\caption{The \texttt{insert} function in give-up template annotated with separation logic specification}
	\label{proof_giveup_insert}
\end{figure}

>>>>

Logical atomicity is key to the template approach: each template's \lstinline{traverse} function is proved to satisfy a logically atomic specification that says roughly ``this function finds the node where key \lstinline{k} belongs''. The \lstinline{traverse} specification can then be used to prove atomic specifications for the individual data structure operations, lifting the sequential specifications for insert, lookup, etc. to the concurrent setting.

@@

Recall that in the lock-coupling pattern, each thread always holds at least one lock while traversing the data structure, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search. Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{n}$ and obtain the resources $\mathsf{R}(\texttt{n})$ guarded by the lock. These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. In formal terms, the resource is represented as $\mathsf{R}(\texttt{n}) \triangleq \mathsf{ghost\_node(\texttt{n}, k, v, range)} \ \ast \ \mathsf{DataStruct\_specific(\texttt{n})}$.

@@

In the specification, the local precondition includes both the node handle $\nodeboxrep(\texttt{n})$ and the contents $\mathsf{R}(\texttt{n})$. The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}. The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a Boolean variable $\mathit{res}$ indicating whether \texttt{n'->t} is \texttt{NULL} or not. The variable $\mathit{res}$ is $\mathsf{true}$ if $\texttt{flag}$ returns 1, and $\mathsf{false}$ if 0, mirroring its Boolean state. The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range}(\texttt{n'})\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \  \mathsf{R}(\texttt{n'}) \ \ast \ \treerep\ m
\end{mathpar} %\wm{Where does \lstinline{range} come from here?}

@@

The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup},  is simpler than that of the lock-coupling template. The local preconditions include any locks, and consist only of $\infp(\texttt{n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$. As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \texttt{n'}'s lock and returning its contents $\mathsf{R}(\texttt{n'})$. Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{n'->t} is \texttt{NULL} or not.

@@

The \texttt{insertOp} specification (Figure \ref{fig:insertOp_giveup}) differs slightly from the lock-coupling version, to account for the added \lstinline{min} and \lstinline{max} fields. In a BST, if the current node's range is \lstinline{(min, max)} and its key is \lstinline{k}, its left child will have range \texttt{(min, k)} and its right child will have range \texttt{(k, max)}. Otherwise, \lstinline{insertOp} behaves identically, storing \lstinline{k} and \lstinline{v} in the current node and creating two empty child nodes.

@@

\begin{figure}[!ht]
	$\color{blue}
	\forall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \infp (\texttt{n}) \ \big| \ \treerep\ m
	\end{array}
	\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \ \texttt{n'}.
	\begin{array}{c}
		\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
		\\ \infp(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{The \texttt{traverse} function in give-up template annotated with separation logic specification}
	\label{proof_giveup_traverse}
\end{figure}

\begin{figure}[!ht]
	\begin{subfigure}{\textwidth}
		$\color{blue}
		\forall m.\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_insert.c}
		$\color{blue}
		\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle$
		\caption{The \texttt{insert} function in lock-coupling template annotated with separation logic specification. \\\\}
		\label{proof_lock_insert}	
	\end{subfigure}\qquad
	\begin{subfigure}{\textwidth}
		$\color{blue}
		\forall m.\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ m
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_insert.c}
		$\color{blue}
		\left\langle 
		\nodeboxrep\ \texttt{r} \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle$
		\caption{The \texttt{insert} function in give-up template annotated with separation logic specification. \\\\}
		\label{proof_giveup_insert}
	\end{subfigure}
	\caption{The \texttt{insert} function in lock-coupling and give-up templates annotated with separation logic specification.}
	\label{proof_lock_giveup_insert}
\end{figure}
